The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-11 kalman_walk_paper.tex
overleaf edits

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals
with a speed depending on the local geometry of the optimal set
(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{1 + (\sigma_S/\gamma)^2} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
We discuss the implications of this expression in the next section.

@@

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
\plr{continue editing here}
(say something about here we interpret the big math expression in words)
From equation \eqref{eqn:fit2} we can observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

argue sometimes $\sigma_N \approx \sigma_S \approx \gamma$, especially if $N_e$ is big
(``large species'')

and that sometimes $\sigma_N \approx \sigma_S \ll \gamma$, but only if $N_e$ is small
(``isolated population'')

and that sometimes maaybe $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load
(``species complex'')

@@

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.

>>>>

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals
with a speed depending on the local geometry of the optimal set
(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\plr{figure out how dimensions come in here}
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{(1 + (\sigma_S/\gamma)^2)} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
We discuss the implications of this expression in the next section.

@@

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
Above, equation \eqref{eqn:fit2} shows how fast hybrids become inviable 
as the time that the parental populations are isolated increases;
what does this tell us about speciation rates under neutrality?
From equation \eqref{eqn:fit2} we can observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
In particular, $\omega$, the rate that hybrid variance increases with separation between parentals,
is proportional to $p-q$, the number of ways mutation can move the system without changing the phenotype,
i.e., the dimensionality of the optimal set.
This implies that less constrained systems should drift faster,
and so lead to speciation more quickly, given the same strength of selection ($\gamma$).
\plr{revisit}
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

% simgaN = sigmaS = gamma, large population
Suppose in a large, genetically diverse population,
the amount of heritable variation in the neutral and selected directions are roughly equal 
($\sigma_N \approx \sigma_S$)
but the overall amount of variation is (weakly) constrained by selection
($\sigma_N \approx \gamma$).
If so, then the first term of equation \eqref{eqn:fit2} is
$1/\sqrt{1 + 2 \omega T/N_e} \approx 1 - \omega T/N_e$.
If also $\omega=1$, then, for instance, after $0.1 N_e$ generations
the average $F_2$ fitness has dropped by 10\% relative to the parentals.

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a^2 T/N_e} \approx 1 - 2 \omega a T/N_e$.
Since in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N$) is proportional to $N_e$,
so this predicts the rate of decrease of hybrid fitness to be \emph{independent} of population size
for small populations.
\plr{look up what $\sigma_N^2$ is at mut-drift equilib}

% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
In the other direction, consider large metapopulations (or isolated ``species complexes'')
among which heritable variation is strongly constrained by selection
(i.e., there is substantial recombination load),
so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a^2 T/N_e} \approx 1 - \omega a T/N_e$,
and could be extremely rapid if $a$ is large.

@@

\begin{figure}
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop if $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N = \sigma_S = \gamma$
\textbf{(B)} $\sigma_N = \sigma_S = 0.1 \gamma$
\textbf{(C)} $0.1 \sigma_N = \sigma_S = \gamma$.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

@@

\jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}
\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
